---
title: "Untitled"
output: html_document
date: "2025-03-12"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      fig.align = "center",
                      #send figures generaed in this file to folder below
                      fig.path = "../figures/03_PreProcessing/")
```

#Goals

Remove any potential contaminants and evaluate the accuracy of our sequencing run. Then, we will write our final data projects

1. load in data that we've generated: `asv_table`,  `tax_table`, and metadata(which includes `metadata.csv` and `track_reads_reads). We created these files in `analysis/02_AssignASV.Rmd
2. Remove any ASVs that are chloroplasts.
3. Remove ASVs that are mitochondria.
4. Evaluate any ASVs from the negative controls. Then, remove negative controls.
5. Evaluate the mock community (positive control). Accuracy of sequencing.
6. Check for reverse complements.
7. Check the sequencing depth of samples. Remove obvious samples with too few reads.


## Input

1. Metadata: `metadata.csv` and `track_read_counts.RData`.
2. ASV table: `data/01_DADA2/ASV_table.csv`.
3. Taxonomy table: `data/01_DADA2/ASV_taxanomy.tsv`

## Output

1. A pre-processed phyloseq object (54 object): `raw_preprocessed-physeq.RData`

# set environment

##load packages
```{r load-packages}
install.packages("BiocManager")
BiocManager::install("Biostrings")

#load packages with pacman
pacman::p_load(devtools, phyloseq, Biostrings, tidyverse, install=FALSE)
```

## load data

### 1. metadata
- copy '/workdir/in_class_data/metadata.csv'
- copy '/workdir/in_class_data/track_read_counts.RData'
- to data folder

asv table=asv (row) x sample names
tax=asv x taxonomy
metadata=samplename x info
```{r load-metadata}
metadata_df<-
 read_csv("data/metadata.csv")%>%
   #fix column names
   rename("sample_names" = "...1" ) %>%
   #add sample names also as a column
   mutate(names=sample_names)

# inspect
head(metadata_df)
dim(metadata_df)

# include dada2 output
load("data/01_DADA2/track_read_counts.RData")

# take a look
head(track_counts_df)
dim(track_counts_df)


# fix sample names and track reads
track_counts_df$sample_names<-sapply(strsplit(track_counts_df$sample_names,"_"),`[`,1)

# intuition check
head(track_counts_df$sample_names)

#what's different?
setdiff(track_counts_df$sample_names, metadata_df$sample_names)

#let's do a filtering join with left join
metadata_final_df<-
  metadata_df%>%
    left_join(.,track_counts_df, by="sample_names")%>%
    # sample names to the rownames to merge into phyloseq
    column_to_rownames(var="sample_names")

#check
dim(metadata_final_df)
```

## 2. ASV table
copy ASV_table.tsv, ASV_taxonomy.tsv data over to my workdir, salinity, data, 01_DADA2
```{r load-asv-table}
asv_df<-
read.delim(file="data/01_DADA2/ASV_table.tsv", sep="\t",
           # add the column names and row names
           header = TRUE, row.names = 1)%>%
  dplyr::select(-"CJ.V08.P")

# inspect
asv_df[1:3,1:3]

#fix column names
##remove the X: denote only X at the beginning "^"
colnames(asv_df)<-sub(pattern="^X", replacement="",colnames(asv_df))
## replace the . with a -: "\\." since . is a special charater in regex
colnames(asv_df)<-gsub(pattern="\\.", replacement="-" , colnames(asv_df))

#find inspection
head(colnames(asv_df))
asv_df[1:3, 1:3]
```

## 3. taxonamy table
```{r load-tax-table}
tax_df<-read.delim("data/01_DADA2/ASV_taxonomy.tsv", sep="\t",
           # add the column names and row names
           header = TRUE, row.names = 1)

dim(tax_df)
dim(asv_df)

#add check
stopifnot(rownames(asv_df)==rownames(tax_df))
```

#hand off to phyloseq
```{r phyloseq-handoff}
raw_physeq<-
phyloseq(otu_table(asv_df,taxa_are_rows = TRUE),
         sample_data(metadata_final_df),
         tax_table(as.matrix(tax_df)))

#save
save(raw_physeq, file="data/03_PreProcessing/raw_physeq.RData")

```


# clean up the data

remove chloroplasts and mitochondria
```{r rm-chloro}
noChloros_physeq<-
raw_physeq%>%
  subset_taxa(Order !="Chloroplast" | is.na(Order))

#how many asvs were chloroplasts?
numChloros_ASVs<-ntaxa(raw_physeq)-ntaxa(noChloros_physeq)
```

what about mitochondria
```{r rm-mitos}
noChlorosMitos_physeq<-
noChloros_physeq%>%
  subset_taxa(Family !="Mitochondria" | is.na(Family))

#how many ASVs were mitochondria?
numMitos_ASVs<-ntaxa(noChloros_physeq)-ntaxa(noChlorosMitos_physeq)

```

# evaluate and remove control samples
```{r neg-controls}
#make a phyloseq object with only the negative controls
neg_control_samples<-c("WaterControl","022um-Control","3um-Control","DNA-Ext-Control")

neg_control_physeq<-
noChlorosMitos_physeq%>%
  subset_samples(., names %in% neg_control_samples)%>%
  # rm ASVs with 0 counts
  prune_taxa(taxa_sums(.)>0,.)

#take a look at the ASV abundances in our negative controls
neg_control_physeq%>%
  plot_bar(., "ASV", fill = "ASV")

#what's the abundance of these neg control asvs in real samples
# create vector of asvs
neg_controlASVs_vec<-
  neg_control_physeq%>%
  tax_table()%>%
  data.frame()%>%
  dplyr::select(ASV)%>%
  as.vector()
  
# use vector to subset from larger dataset, creating a new phyloseq object
neg_controlASVs_physeq<-
noChlorosMitos_physeq%>%
  subset_taxa(., ASV%in% neg_controlASVs_vec$ASV)

#plot abundances of neg controls ASVs
neg_controlASVs_physeq%>%
  plot_bar(.,"ASV", facet_grid = Sample_or_Control~.)

#create a new vector with all ASVs except ASV17 and 50
negControlASVs_rm<-dplyr::setdiff(neg_controlASVs_vec$ASV, c("ASV_0017","ASV_0050"))

#make new phyloseq object with neg control asvs removes
noChloroMitosNegControls_physeq<-
noChlorosMitos_physeq%>%
  #rm ASVs
  subset_taxa(., !(ASV %in% negControlASVs_rm))%>%
  #rm neg control samples
  subset_samples(.,!(names %in% neg_control_samples))
```





