---
title: "04A_Phylogenetics_Tree_Construction"
output: html_document
date: "2025-03-19"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Goals
the goal is to make the phylogenetic tree

1. load in our pre-processed phyloseq object (raw_preprocess_phyloseq.RData)
2. create ASV fasta file from the phyloseq object
3. align the 16s sequences from our fasta file with MAFT
4. create a maximum likelihood tree using FastTree2


### Input
1. raw_preprocessed_physeq.RData

### Output
1. MAFT_aligned_ASV.fasta
2. ASVs_unrooted.tree

# set seed
```{r set-seed}
set.seed(20020718)
```

# load packages and phyloseq project
```{r load-environment}

# install phytools
remotes::install_github("liamrevell/phytools")

#install ggtree
devtools::install_github("GuangchuangYu/ggtree")

pacman::p_load(phytools, ggtree, RColorBrewer, phyloseq, install = FALSE)

# load physeq
load("data/03_PreProcessing/raw_preprocessed_physeq.RData")

raw_preprocessed_physeq
```
# Write ASV fasta file
```{r create-preprocessed-fasta}
# pull out ASV sequences and ASV names
asv_seq_df<-
  raw_preprocessed_physeq@tax_table %>%
  data.frame() %>% 
  dplyr::select(ASV, ASVseqs)

#View(asv_seq_df)

#add the ">" to make the fasta header
asv_seq_df$ASV <- paste0(">", asv_seq_df$ASV)

#create fasta object
asv_seq_fasta<- c(rbind(asv_seq_df$ASV,asv_seq_df$ASVseqs))

#write to a fasta file
write(asv_seq_fasta,
      file="data/04_Phylogenetic_Tree/preprocessed_ASVs.fasta")
```

# Align the 16s sequences from fasta file with MAFT
- engine.opts='-l': this gives us our original terminal base environment

to test shell directly from rmd: mac: cmd_option_enter
```{r run-maft, engine='bash', engine.opts='-l'}
# Write bash code to run MAFT alignment

# first provide the path
#highlight then cmd_option_enter
export PATH=/programs/mafft/bin:$PATH

#change directories to provide the fasta file we made above
cd data/04_Phylogenetic_Tree

pwd

#set a seed for consistency and reproducibility

RANDOM=20020718

#MAFFT already knows we have a nucleotide alignment in fasta file
#this line is found in cornell biohpc-software-MAFFT
/programs/mafft/bin/mafft --auto preprocessed_ASVs.fasta > MAFFT_aligned_ASVs.fasta

```


# FastTree2
```{r run-fasttree, engine='bash', engine.opts='-l'}
# Provide export path to fasttree
export PATH=/programs/FastTree-2.1.11:$PATH

#change directory to alignment file folder
#cd data/04_Phylogenetic_Tree/

# run fasttree to generate phylogenetic tree
# parameters
  #nt: indicate it's a nucleotide alignment
  #gtr: generalized time reversible substitution model
  #fastest: speed up the model, reduce memory usage, recommended for dataset with>50000 sequences
  #log: output a log file
  #input our alignment file
  #specify the output tree file
FastTree -nt -gtr -fastest -log FastTree.log MAFFT_aligned_ASVs.fasta > ASVs_unrooted.tree

# change back to main directoy

cd ../../

echo "The working directoy is $PWD"
```

