---
title: "04B_Phylogenetics"
output: html_document
date: "2025-03-19"
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      fig.align = "center",
                      fig.path = "../figures/04_Phylogenetic_Tre")
```

#Goal: 
investing and root our tree

1. load the FastTree unrooted tree
2. add the tree to phyloseq object
3. visualize tand inspect tree with ggtree
4. prune ASVs, if needed
5. root our tree
6. combine newly rooted tree with a phyloseq object
7. save 2 phylotree object
  1. Unrooted tree phyloseq object
  2. Rooted tree phyloseq object
  
# Input
1. raw_preprocessed_physeq.RData
2. ASVs_unrooted.tree
3. MAFFT_aligned_ASVs.fasta

# output
1. phytree_preprocessed_physeq.RData

#set my seed
```{r set-seed}
set.seed(20020718)
```

## load packages
```{r load-packages}
pacman::p_load(tidyverse, phyloseq, ggtree, phytools, tidytree,
               install = FALSE)
```


## load data files
```{r load-data}
# Preprocessed phyloseq object 
load("data/03_PreProcessing/raw_preprocessed_physeq.RData")
raw_preprocessed_physeq

# Load in the tree! 
unrooted_tree <- read.tree("data/04_Phylogenetic_Tree/ASVs_unrooted.tree")
unrooted_tree

# view structure of unrooted_tree
str(unrooted_tree)
```

# merge phyloseq object
```{r merge-physeq}
# Intuition check 
stopifnot(ntaxa(raw_preprocessed_physeq) == ntaxa(unrooted_tree))

# Merge the tree with the phyloseq object 
unrooted_physeq <- 
  merge_phyloseq(raw_preprocessed_physeq, unrooted_tree)

unrooted_physeq
```

# plot tree with `ggtree`
```{r plot-tree-unrooted}
#make a basic tree
kingdom_tree <- 
  ggtree(unrooted_physeq) + 
  # color tips by kingdom 
  geom_tippoint(mapping = aes(color = Kingdom)) + 
#  scale_color_manual(values = c("goldenrod1", "cornflowerblue", "grey")) +
  # Add title 
  labs(title = "Unrooted Tree") + 
  #move the legend to the bottom 
  theme(legend.position = "bottom"); kingdom_tree

kingdom_tree

# view tree with labels
kingdom_node_tree<-
  kingdom_tree +
  #add node label
  geom_text(aes(label=node), hjust=-0.5, vjust=-0.3,size=1)
kingdom_node_tree
```

#root tree from archea
go to line 232 if no archea in dataset so can midroot tree
```{r root-archea}
# first find the ASVs that are Archaea
archaea_node <- tax_table(raw_preprocessed_physeq) %>%
  as.data.frame() %>%
  filter(Kingdom == "Archaea") %>%
  pull(ASV)

# find the node that encompasses all Archaea
archaea_node <- findMRCA(unrooted_tree, tips = archaea_node, type = "node") 
# gives node = 3394

# View clade to check it out
viewClade(kingdom_node_tree, archaea_node)

# root the tree
node3394<-
  ggtree(unrooted_physeq)+
  geom_tippoint(mapping= aes(color=Kingdom),
                size = 1.5) +
  geom_text(aes(label = node), hjust = -0.7, vjust = -0.2, size = 2) +
  geom_highlight(node = 3394, fill = "dodgerblue")

node3394

#root the tree using 3394
archaea_root_tree <- ape::root(unrooted_tree, node = archaea_node, resolve.root = TRUE)

# Check if tree is rooted
is.rooted(archaea_root_tree) # should be TRUE

# merge with physeq
archaea_root_physeq <- merge_phyloseq(raw_preprocessed_physeq, archaea_root_tree)

#plot rooted simple tree with no text
archae_rooted_simple<-
  ggtree(archaea_root_physeq)+
  geom_tippoint(aes(color = Kingdom))  

#plot rooted tree with text
archaea_rooted_text<-
  ggtree(archaea_root_physeq)+
  geom_tippoint(aes(color = Kingdom)) + 
  geom_text(aes(label = node), hjust = -0.5, vjust = -0.3, size = 1.5)

```

#evaluate long branches
```{r evaluate-long-branches}
# View a specific clade

# Zoom in on node that looks potentially suspicious
viewClade(archaea_rooted_text + 
          labs(title = "Archaea Rooted Tree: Node 703"), 
          node = 703)

#what are its ancestors
ancestor(archaea_root_tree, 703)

# Or if there are two nodes that we can see we can find its MRCA

MRCA(archaea_root_tree, .node1 = 703, .node2 = 702) #3309
```

