---
title: "Biodiversity across a salinity gradient"
output: html_document
date: "2025-03-26"
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      fig.align = "center",
                      fig.path = "../figures/05_Biodiversity")
```

# Goals:

Run an analysis on the within-sample biodiversity of the samples in out project

1. calculate the Hill Numbers of the samples using the iNEXT package
2. Plot and evaluate the interpolated and extrapolated rarefaction curves from iNEXT
3. analyze the hill numbers as it related to stations and salinity (in PSU)
4. answer the scientific question on hand!

## Input:

1.`phytree_preprocessed_physeqs`

## Output:
1. calcualte biodiversity measures (hill numbers) across every sample
2. figures to visualize these changes in biodiversity
3. statistical tests to actually show the changes and patterns in biodiversity

# set up the environment


```{r set-seed}
set.seed(20020718)
```

## load packages and colors

```{r load-packages-colors}
pacman::p_load(tidyverse, devtools, patchwork, iNEXT, phyloseq, 
               install = FALSE)

source("code/colors.R")
```

```{r load-data}
# Preprocessed phyloseq object: With the phylogenetic tree!
load("data/04_Phylogenetic_Tree/phytree_preprocessed_physeqs.RData")

archaeal_rooted_physeq

#make a metadata dataframe
metadata_df <-
  archaeal_rooted_physeq %>%
  sample_data() %>%
  data.frame()

# Take a look!
head(metadata_df)
```

#calcualte biodiversity with iNEXT
```{r iNEXT-biodiversity}
#prepare the input datda
##species in rows nd samples in columns
iNEXT_input_df <- 
  archaeal_rooted_physeq %>%
  otu_table() %>%
  data.frame()

# Quick check
dim(iNEXT_input_df)
iNEXT_input_df[1:5,1:5]

#run iNEXT
#remember to set the seed
#ASVs in rows, samples in columns
iNEXT_data<-iNEXT(iNEXT_input_df,
                  Q=C(0,1,2),
                  datatype = "abundance")

```

since iNEXT takes a while to run, we'll load in calculated data
```{r load-iNEXT-data}
load("data/05_Biodiversity/iNEXT_data.RData")

#inspect
str(iNEXT_data)
typeof(iNEXT_data)
```

#plot the rarefaction curves

```{r iNEXT-colors}
# Prepare Colors 
color_df <- 
  iNEXT_input_df %>%
  colnames() %>%
  data.frame()
# Check
head(color_df)
# Rename the column 
colnames(color_df)[1] <- "names"
# Check
head(color_df)


# Make a helper dataframe for plotting with colors 
iNEXT_color_df <- 
  color_df %>%
  # Fix the names for merging
  mutate(names = gsub(names, pattern = "[.]", replace = "-"),
         names = gsub(names, pattern = "X",  replace = "")) %>%
  # Merge with metadata
  left_join(metadata_df, by = "names") %>%
  # Merge with colors for plotting with ggiNEXT
  left_join(data.frame(station_colors = station_colors,
            station = names(station_colors)),
            by = "station")

head(iNEXT_color_df)
```

remember the output of iNEXT
q=0, richness
q=1, shannon
q=2, simpson
```{r plot-iNEXT-rarefaction}
ggiNEXT(iNEXT_data,type=1,facet.var="Order.q")+
  scale_color_manual(values = iNEXT_color_df$station_colors, guide = FALSE) + 
  scale_fill_manual(values = iNEXT_color_df$station_colors, guide = FALSE) +
  scale_shape_manual(values = base::rep(17, nsamples(archaeal_rooted_physeq)),
                     guide = FALSE) +
  labs(x = "Number of Sequences (Library Size)", 
       y = "Effective Number of ASVs") + 
  theme_bw() + 
  theme(legend.position = "none") 


#big drop between q=0 and q=1 plots represent the long tail

```

manually plot rarefaction

```{r manual-rarefaction}
str(iNEXT_data)


iNEXT_manual_df <- 
  iNEXT_data$iNextEst$size_based %>%
  dplyr::rename(names = Assemblage) %>%
  # Fix the samples names 
  mutate(names = gsub(names, pattern = "[.]", replace = "-"),
         names = gsub(names, pattern = "X", replace = "")) %>%
  # join with metadata 
  left_join(., metadata_df, by = "names") %>%
  # Add colors to data frame
  left_join(., data.frame(station_colors = station_colors,
                          station = names(station_colors)),
            by = "station") %>%
  mutate(station = factor(station, levels = c("Copano West", "Copano East", 
                                              "Mesquite Bay", "Aransas Bay", 
                                              "Shipping Channel")))
# Inspect 
dim(iNEXT_manual_df)

# manually plot the rarefaction curves
iNEXT_manual_df %>%
  # Filter out rows that are calculated with actual sequences (no extrapolation)
  dplyr::filter(Method == "Rarefaction") %>%
  # Now, let's draw the plot, be sure to group by the samples if using geom_line!
  ggplot(aes(x = m, y= qD, color = station, group = names)) + 
  geom_line() + 
  # Facet with the station to see the samples better 
  facet_grid(Order.q~station, scales = "free") + 
  scale_color_manual(values = station_colors) + 
  labs(x = "Number of Sequences (Library Size)", 
       y = "Effective Number of ASVs") + 
  theme_bw() + 
  #guides(color = guide_legend(nrow = 2, byrow = TRUE)) + 
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 30, vjust = 1, hjust = 1),
        legend.title = element_blank()) 

```

#boxplot: diversity by station
## categorical analysis
```{r make-obs-df}
# Make a dataframe
obs_div_df <- 
  iNEXT_manual_df %>%
  dplyr::filter(Method == "Observed") %>%
  left_join(data.frame(station_names = station_names, 
                       station = names(station_names)), 
                       by = "station") %>%
  mutate(station_names = factor(station_names, 
                                levels = c("Copano\nWest", "Copano\nEast", 
                                           "Mesquite\nBay", "Aransas\nBay", 
                                           "Shipping\nChannel")))

# Check it 
head(obs_div_df)

#plot boxplot by station against diversity
obs_div_df %>%
  ggplot(aes(x = station_names, y = qD, fill = station, color = station)) + 
  facet_wrap(.~Order.q, scales = "free_y") + 
  geom_jitter(size = 2.5) + 
  geom_boxplot(alpha = 0.5, outlier.shape = NA) + 
  scale_color_manual(values = station_colors) + 
  scale_fill_manual(values = station_colors) + 
  labs(y = "Effective Number of ASVs") + 
  theme_bw() +
  theme(legend.position = "bottom",
        axis.title.x = element_blank(),
        axis.text.x = element_text(angle=30,
                                   hjust=1,
                                   vjust=1),
        legend.title = element_blank())

```


# continuous relationships - salinity(PSU) vs biodiversity

## environmental variables
```{r environ-check}
env_df<-
  metadata_df%>%
  dplyr::select(names, water_tempC:DO_mgL)

#pair-wise interactions between env var
pairs(dplyr::select(env_df, -names), upper.panel = NULL)
```

## PSU
-practical salinity unit: 1 PSU=1g salt/1000g water =1 ppt
-gradient typically 0->30
-freshwater: 0-0.5
-brackish water: 0.5-30
-marineL:>30

```{r plot-div-salinity-lm}
#how does biodiversity change over these salinity gradients
obs_div_df%>%
  ggplot(aes(x = salinity_psu, y = qD)) + 
  facet_wrap(.~Order.q, scales = "free") + 
  geom_point(aes(color = station)) + 
  stat_smooth(method = "lm", formula = y ~poly(x, 2)) + 
  labs(x = "Salinity (PSU)", y = "# of ASVs") + 
  scale_color_manual(values = station_colors) + 
  theme_bw() + 
  theme(legend.position = "bottom",
        legend.title = element_blank()) 
```

