---
title: "Between-Sample (Beta) Diversity of Microbes along a Salinity Gradient"
output: html_document
date: "2025-04-09"
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      fig.path = "../figures/06_CommunityAnalysis")
```

# Goal

1. load in phyloseq data with rooted tree, colors, functions.
2. evaluate the sequencing depth and, if needed, remove samples.
3. normalize the read counts between samples.
4. calculate community **dis**similarities: numbers will always mean 0=same,1=dissimilar/different.
    - Sorensen
    - Bray-Curtis
    - UniFrac (weighted & unweighted)
5. run some statistics: PERMANOVA & betadisper
6. plot some bar plots of specific microbial groups in the data

## Input

1. phylogenetic tree phyloseq object that we created in 'analysis/04_phylogenetic_tree_inspection.Rmd`.

## Output

1. calculated beta-diversity dissimilarity measures.
2. figures: PCoA and NMDs
3. statistical tests: PERMANOVA & betadisper

# scientific question

How does microbial concentration change across a salinity gradient?

- *null hypothesis*: no change in microbial community iwth varying salinity
- *alternative hypothesis*: community composition will change with varying salt concentrations
  - rare members? jaccard, **sorensen**, unweighted unifrac
  - dominant members? **bray-curtis**, weighted unifrac
  - evolutionary related? unweighted and weighted unifrac
  

# set up
## set seed
```{r set-seed}
set.seed(20020718)
```

## load packages
```{r load-packages}
pacman::p_load(tidyverse, devtools, phyloseq, patchwork, vegan, install = FALSE)

#load colors
source("code/colors.R")
#load functions
source("code/functions.R")
```

# 1. load in the data
```{r}
load("data/04_Phylogenetic_Tree/phytree_preprocessed_physeqs.RData")
archaeal_rooted_physeq

#intuition check on seq depth
min(sample_sums(archaeal_rooted_physeq))

#create metadata dataframe
meta_df<-
  archaeal_rooted_physeq %>%
  sample_data()%>%
  data.frame()
```

# normalizing the read depth

## 2. explore the raw read counts
```{r explore-read-counts}
# calculate read counts per sample
raw_TotalSeqs_df<-
  archaeal_rooted_physeq%>%
    #calculate the total number of sequences/reads
  sample_sums()%>%
  data.frame()

# take a look
head(raw_TotalSeqs_df)

# rename the column
colnames(raw_TotalSeqs_df)[1]<-"TotalSeqs"

# add a new column of num_ASVs (RAW, non-normalized #ASVs)
raw_TotalSeqsASVs_df<-
  raw_TotalSeqs_df%>%
  mutate(num_ASVs = colSums(otu_table(archaeal_rooted_physeq)>1))

#view(raw_TotalSeqsASVs_df)

# plot a hisotgram of seq depth 
raw_TotalSeqsASVs_df%>%
  ggplot(aes(x=TotalSeqs))+
  geom_histogram(bins=50)+
  #check max in console max(raw_TotalSeqsASVs_df$)
  scale_x_continuous(limits =c(0,7500) )+
  labs(title = "Raw Seq Depth")+
  theme_bw()

# plot seq depth vs num_ASVs
raw_TotalSeqsASVs_df%>%
  ggplot(aes(x=num_ASVs, y=TotalSeqs))+
  geom_point()+
  scale_y_continuous(limits=c(0,7500))+
  geom_smooth(method="lm")+
  labs(title = "Seq depth vs # ASVs")

#check linear relationship
summary(lm(TotalSeqs=num_ASVs, data=raw_TotalSeqsASVs_df))

```

# 3. scaling read counts
```{r scale-reads}
min(sample_sums(archaeal_rooted_physeq))

#scale the reads
scaled_physeq<-
  archaeal_rooted_physeq%>%
  scale_reads(round="matround")

#look at it
scaled_physeq

# confirm the sequencing read depth of scaled_physeq
scaled_TotalSeqs_df<-
  scaled_physeq%>%
  sample_sums()%>%
  data.frame()

colnames(scaled_TotalSeqs_df)[1]<-"TotalSeqs"

# plot it!
scaled_TotalSeqs_df%>%
  ggplot(aes(x=TotalSeqs))+
  geom_histogram(bins=50)+
  scale_x_continuous(limits=c(0,7500))+
  theme_bw()

# scaling always gives us a range!
min(scaled_TotalSeqs_df$TotalSeqs)
range(scaled_TotalSeqs_df)
```

# calculate & visualize community dissimilarity

two different visualizations:
1. PCoA - Eigen-based analysis, linear, axis coefficients are meaningful because they represent the eigenvectors
2. NMDS - non-linear method, axes are not meaningful, shmush many axes of data into 2 axes

## sorensen dissimilarity 

### calculate dissimilarity
```{r soren-dist}
scaled_soren_dist<-
  phyloseq::distance(scaled_physeq, method="bray",binary=TRUE)

class(scaled_soren_dist)
#distance matrix: sample to sample

## PERMANOVA: are centroids of clusters different?
adonis2(scaled_soren_dist ~ station, data=meta_df)
adonis2(scaled_soren_dist ~ station+ date, data=meta_df, by="margin")
adonis2(scaled_soren_dist ~ station* date, data=meta_df, by="margin")

#betadisper
dispr_soren_station<-betadisper(scaled_soren_dist,meta_df$station)
permutest(dispr_soren_station)
```

### plot PCoA
```{r soren-pcoa}
# first, calculate pcoa with sorenson
scaled_soren_pcoa<-
  ordinate(physeq = scaled_physeq,
           method = "PCoA",
           distance = "bray", binary=TRUE)

#take a quick look
str(scaled_soren_pcoa)

# plot it!
plot_ordination(physeq = scaled_physeq,
                ordination = scaled_soren_pcoa,
                color="station",
                title = "Sorensen PCoA")+
  scale_color_manual(values=station_colors)+
  geom_point(size=5, alpha=0.5,aes(color=station))
  theme_bw()+
  theme(legend.position = "bottom")
  
# u shaped, not good, maybe need more than two dimensions
  
  
```


### plot NMDS
```{r soren-nmds}
scaled_soren_nmds<-
  ordinate(physeq = scaled_physeq,
           method = "NMDS",
           distance = "bray", binary=TRUE)

plot_ordination(physeq = scaled_physeq,
                ordination = scaled_soren_nmds,
                color="station",
                title = "Sorensen NMDS")+
  scale_color_manual(values=station_colors)+
  geom_point(size=5, alpha=0.5,aes(color=station))
  theme_bw()+
  theme(legend.position = "bottom")
  
  #right now close points mean having similar taxa, particularly rare, because binary measure
  # if bray-curtis, two points close have similar dominant taxa

  #want stress level to be 0.15
```

